% Copyright 2018 Beckman Coulter, Inc.
%
% Permission is hereby granted, free of charge, to any person
% obtaining a copy of this software and associated documentation files
% (the "Software"), to deal in the Software without restriction,
% including without limitation the rights to use, copy, modify, merge,
% publish, distribute, sublicense, and/or sell copies of the Software,
% and to permit persons to whom the Software is furnished to do so,
% subject to the following conditions:
%
% The above copyright notice and this permission notice shall be
% included in all copies or substantial portions of the Software.
%
% THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
% EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
% MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
% NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
% BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
% ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
% CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
% SOFTWARE.

\documentclass[letterpaper,11pt,twoside,final]{article}
\usepackage{sagian}

\begin{document}

\title {C Coding Standard}
\author {Bob Burger}
\date {\copyright\ 2018 Beckman Coulter, Inc.
  Licensed under the \href{https://opensource.org/licenses/MIT}{MIT License}.}
\coverpage

\section* {Identifiers}

Use meaningful names---this can be very difficult. Identifiers should
correspond directly to the names used in the design
documents. Identifiers should use lower-case letters with an
underscore separating compound words. Mixed-case identifiers may be
used when they directly correspond to an external definition.

\section* {Comments}

Comments should be clear, concise, and written in U.S. English.

Avoid using comments that say what the code says.

\section* {Formatting}

C code should be indented with Artistic
Style\footnote{\url{http://astyle.sourceforge.net}} configured with a
project-specific \code{.astylerc} file.

Line length should be limited to 79 characters. Lines longer than 79
characters may be tolerated.

Lines should end in LF, not CRLF.

Tab characters should not appear in C code. Their presence in a file
makes the indentation depend on the software used to view the file.

\end{document}
