% Copyright 2018 Beckman Coulter, Inc.
%
% Permission is hereby granted, free of charge, to any person
% obtaining a copy of this software and associated documentation files
% (the "Software"), to deal in the Software without restriction,
% including without limitation the rights to use, copy, modify, merge,
% publish, distribute, sublicense, and/or sell copies of the Software,
% and to permit persons to whom the Software is furnished to do so,
% subject to the following conditions:
%
% The above copyright notice and this permission notice shall be
% included in all copies or substantial portions of the Software.
%
% THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
% EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
% MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
% NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
% BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
% ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
% CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
% SOFTWARE.

\documentclass[letterpaper,11pt,twoside,final]{article}
\usepackage{swish}

\begin{document}

\title {Scheme Testing Library}
\author {Leif Danielson \and Bob Burger}
\date {\copyright\ 2018 Beckman Coulter, Inc.
  Licensed under the \href{https://opensource.org/licenses/MIT}{MIT License}.}
\maketitle

\section* {Introduction}

The Scheme Testing Library \code{(swish mat)} provides methods to
define, iterate through, and run test cases, and to log the
results. Test cases are called \emph{mats} and consist of a name, a
set of tags, and a test procedure.  The set of mats is stored in
reverse order in a single, global list.  The list of tags allows the
user to group tests or mark them.  For example, tags can be used to
note that a test was created for a particular change request.

The Scheme Testing Library uses only the \code{(chezscheme)}
library so that it can be used to test any Scheme code.

A \emph{mat} is an entry in the global list of the form
\code{(\var{name} \var{tags} . \var{test})}. The \var{test} is a
procedure of no arguments.

\section* {Programming Interface}

% ----------------------------------------------------------------------------
\begin{syntax}
  \code{(mat \var{name} (\var{tag} \etc) $e_1$ $e_2$ \etc)}
\end{syntax}
\expandsto{} \code{(add-mat '\var{name} '(\var{tag} \etc)
  (lambda () $e_1$ $e_2$ \etc))}

The \code{mat} macro creates a mat with the given \var{name},
\var{tag}s, and test procedure $e_1$ $e_2$ \etc\ using the
\code{add-mat} procedure.

% ----------------------------------------------------------------------------
\begin{procedure}
  \code{(add-mat \var{name} \var{tags} \var{test})}
\end{procedure}
\returns{} unspecified

The \code{add-mat} procedure adds a mat to the front of the global
list. \var{name} is a symbol, \var{tags} is a list, and \var{test} is
a procedure of no argument.

If \var{name} is already used, an exception is raised.

% ----------------------------------------------------------------------------
\begin{procedure}
  \code{(run-mat \var{name} \var{reporter})}
\end{procedure}
\returns{} see below

The \code{run-mat} procedure runs the mat of the given \var{name} by
executing its test procedure with an altered exception handler. If the
test procedure completes without raising an exception, the mat result
is \code{pass}. If the test procedure raises exception \var{e}, the
mat result is \code{(fail~.~\var{e})}.

After the mat completes, the \code{run-mat} procedure tail calls
\code{(\var{reporter} \var{name} \var{tags} \var{result} \var{statistics})}.

If no mat with the given \var{name} exists, an exception is raised.

% ----------------------------------------------------------------------------
\begin{syntax}
  \code{(run-mats)}
\end{syntax}
\begin{syntax}
  \code{(run-mats \var{name$_1$} \var{name$_2$} \etc)}
\end{syntax}
\returns{} unspecified

The \code{run-mats} macro runs each mat specified by symbols
\var{name$_1$} \var{name$_2$} \etc.  When no names are supplied, all
mats are executed.  After each mat is executed, its result, name, and
exception if it failed are displayed.  When the mats are finished, a
summary of the number run, passed, and failed is displayed.

% ----------------------------------------------------------------------------
\begin{procedure}
  \code{(run-mats-to-file \var{filename})}
\end{procedure}
\returns{} unspecified

The \code{run-mats-to-file} procedure executes all mats and writes
the results into the file specified by the string \var{filename}. If
the files exists, its contents are overwritten. The file format is a
sequence of s-expressions readable with \code{load-results} and
\code{summarize}.

% ----------------------------------------------------------------------------
\begin{procedure}
  \code{(for-each-mat \var{procedure})}
\end{procedure}
\returns{} unspecified

The \code{for-each-mat} procedure calls \code{(\var{procedure}
  \var{name} \var{tags})} for each mat, in no particular order.

% ----------------------------------------------------------------------------
\begin{procedure}
  \code{(load-results \var{filename})}
\end{procedure}
\returns{} a Scheme record

The \code{load-results} procedure reads the contents of the file
specified by string \var{filename} and returns a Scheme record:
\codebegin
(define-record-type mat-data
  (nongenerative)
  (fields
    (immutable report-file)
    (immutable meta-data)
    (immutable results)))
\codeend
containing the \var{filename}, a list of meta-data records:
\codebegin
(define-record-type meta-data
  (nongenerative)
  (fields
   (immutable test-file)
   (immutable key)
   (immutable value)))
\codeend
and a list of result records:
\codebegin
(define-record-type mat-result
  (nongenerative)
  (fields
   (immutable test-file)
   (immutable test)
   (immutable tags)
   (immutable type)
   (immutable message)
   (immutable sstats)))
\codeend

% ----------------------------------------------------------------------------
\begin{procedure}
  \code{(summarize \var{files})}
\end{procedure}
\returns{} three values: the number of passing mats, the number of
failing mats, and the number of skipped mats

The \code{summarize} procedure reads the contents of each file in
\var{files}, a list of string filenames, and returns the number of
passing mats, the number of failing mats, and the number of skipped mats.
An error is raised if any
entry is malformed.

\end{document}
