% Copyright 2019 Beckman Coulter, Inc.
%
% Permission is hereby granted, free of charge, to any person
% obtaining a copy of this software and associated documentation files
% (the "Software"), to deal in the Software without restriction,
% including without limitation the rights to use, copy, modify, merge,
% publish, distribute, sublicense, and/or sell copies of the Software,
% and to permit persons to whom the Software is furnished to do so,
% subject to the following conditions:
%
% The above copyright notice and this permission notice shall be
% included in all copies or substantial portions of the Software.
%
% THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
% EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
% MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
% NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
% BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
% ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
% CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
% SOFTWARE.

\chapter {Developing Software with Swish}\label{chap:app}

\section {Introduction}

Swish can be used to build, test, and deploy programs ranging
from small scripts to large stand-alone applications.
This chapter describes some of the tools and mechanisms that Swish provides
for these purposes.

% \section {Theory of Operation}
% perhaps describe boot process?

\section {Deployment Types}

For interactive development, Swish provides a REPL that reads, evaluates,
and prints the values of programs entered at the prompt.
At the REPL, the \code{load} procedure can be used to evaluate the
contents of a file containing source or object code.
This is convenient when developing larger programs.

Swish provides several options for deploying programs.
This section describes these options and their trade-offs.

\subsection {Scripts}

A simple deployment option is to place source code in a file that begins with
a \code{\#!} line specifying the absolute path to an executable that can
evaluate the script.
This could be the absolute path to the Swish executable.
More often \code{/usr/bin/env} is used to locate the Swish executable via the
program search path.
For example, we might have:
\codebegin
$ cat hello
#!/usr/bin/env swish
(printf "Hello, World!{\textbackslash}n")
$ chmod +x hello
$ ./hello
Hello, World!
\codeend

In the preceding example, running \code{./hello} invokes the Swish executable
with \code{./hello} as its sole command-line argument.
At boot time, Swish calls \code{swish-start} to process its command-line arguments.
Since the first argument is not an option (\sopt{h}, \lopt{help}, etc.),
\code{swish-start} runs the file named by that argument with the
remaining arguments, if any, as its command-line arguments.
For example, the following \code{Echo} script processes the arguments
that are passed to the script.

\codebegin\fixtilde
$ cat Echo
#!/usr/bin/env swish
(printf "~\{~:(~a~)~\}{\textbackslash}n" (command-line-arguments))
$ chmod +x Echo
$ ./Echo some camel case identifiers are hard to read
SomeCamelCaseIdentifiersAreHardToRead
\codeend

To provide arguments to the Swish executable before the script filename, add
the \sopt{S} option to \code{env} and add the desired arguments after
\code{swish}.
Here the \lopt{} option tells Swish to start a REPL after loading the script.
\codebegin
$ cat howdy
#!/usr/bin/env -S swish \lopt{}
(printf "Howdy, Folks!{\textbackslash}n")
(pretty-print (command-line-arguments))
$ chmod +x howdy
$ ./howdy

Swish Version 0.0.0
Howdy, Folks!
("--" "./howdy")
> (+ 2 3)
5
> (exit)
\codeend

% to show boot search while loading script:
% #!/usr/bin/env -S swish --verbose

% Note that we *can* write a script named --help, since
% the argument comes in as something like ./--help.

\subsubsection {Limitations}\label{sec:script-limitations}

There are several constraints to consider when deploying Swish scripts. 
\begin{itemize}
  \item
    Naturally, Swish must be installed.
    To use \code{/usr/bin/env} as in the preceding examples, the \code{PATH}
    environment variable must contain the directory where the Swish executable
    is installed.
    This is preferable to hard-coding the absolute path to the Swish executable
    in your scripts.
  \item
    Chez Scheme must also be installed.
    In particular, the version of Chez Scheme that was used to build Swish
    must be installed.
    Swish must be able to locate the Chez Scheme boot files \code{petite.boot}
    and \code{scheme.boot}.
    % (Yes, we currently expect both boot files.)

    If Chez Scheme is installed in a non-standard location, it may be
    necessary to set the \code{SCHEMEHEAPDIRS} environment variable
    to help Swish locate the boot files.
    To see where Swish looks for boot files, run \code{swish \lopt{verbose}}.
  \item
    The \code{\#!} scripts shown in this section do not run under Windows.
    In MinGW/MSYS, these scripts may work unless Posix path conversion is
    disabled by setting the \code{MSYS\_NO\_PATHCONV} environment variable to 1.
    In Cygwin, these scripts may work if the directory containing the script
    is mounted as its Windows twin, e.g., \code{C:/Users} and \code{/Users}.
    For either of these options to work, Swish must be able to locate the
    appropriate Chez Scheme DLL via the standard search order.
    Scripts will fail if the current drive is not the drive containing the script.
\end{itemize}

\subsection {Linked Programs}

A linked program is simply a Scheme object file that begins
with a ``\code{\#!/usr/bin/env swish}'' line.
Swish runs these programs the same way it runs scripts,
except that \code{swish-start} skips the compilation step
when it runs the file.

We can use \code{swish-build} to build a linked program \code{foo} from a
source file \code{foo.ss}, as follows:
\codebegin
$ swish-build -o foo foo.ss
\codeend

\subsubsection {Limitations}

All of the limitations in Section~\ref{sec:script-limitations} apply.
Since linked programs use \code{/usr/bin/env}, the Swish executable
must be in a directory in the \code{PATH} on the target machine.

In addition, the Swish executable (and supporting code) that is used
to run a linked program must be identical to the one that was used to
build it.

\subsection {Stand-alone Programs}

A stand-alone program consists of an executable and a boot file
that must be installed in the same directory.
The name of the boot file is the same as the executable's name
with any extension replaced by \code{.boot}.
The executable is simply a copy of the Swish executable.
The boot file includes the necessary Chez Scheme boot files
along with the compiled application code.

We can use \code{swish-build} to build a linked program \code{foo} from a
source file \code{foo.ss} by specifying a base boot file via the \sopt{b}
option:
\codebegin
$ swish-build -o foo foo.ss -b petite
\codeend

\subsubsection {Limitations}

On Windows, the Swish DLLs \code{osi.dll}, \code{libuv.dll}, and \code{sqlite3.dll}
must be installed in the same directory as the program executable.
These DLLs can be found in the directory containing the Swish executable.
In addition, the program executable must be able to locate
Microsoft's C Runtime Library \code{vcruntime140.dll}
and the appropriate Chez Scheme DLL via the standard search order.

\section{Running Tests}

Use the \code{mat} and \code{isolate-mat} forms described in
Section~\ref{sec:mat} to define automated tests.
Use the \code{swish-test} script to run tests and report the results.
This script treats each file with a \code{.ms} extension as a suite of tests.
See \code{swish-test \lopt{help} all} for details.

% TODO eventually describe
%  - #!eof mats

\section {Programming Interface}

\subsection{Configuration}

\defineentry{app:config}
\begin{procedure}
  \code{(app:config \opt{\var{obj}})}
\end{procedure}
\returns{} a hashtable

When called with no arguments, the \code{app:config} procedure
returns the configuration data cached in a private process parameter.
If the cache is empty, \code{app:config} first populates the cache by reading
data from the file identified by \code{app:config-filename}.
If that file does not exist, the procedure returns an empty hashtable.
Otherwise, \code{app:config} expects the file to contain a single JSON object.

The optional \var{obj} must be \code{\#f} or a hashtable in the form returned by
\code{json:make-object}.
Calling \code{app:config} with \code{\#f} clears the cache.
Calling \code{app:config} with a hashtable installs the hashtable as the cached value.

\defineentry{app:config-filename}
\begin{procedure}
  \code{(app:config-filename)}
\end{procedure}
\returns{} the name of the application's configuration file

The \code{app:config-filename} procedure returns the name of the application
configuration file that \code{app:config} will read.
The filename returned depends on the value of \code{app:path}.

If \code{app:path} is set to a value of the form
\code{"\var{...}/bin/\var{app-name}\opt{.\var{ext}}"},
that ends with a \code{"bin"} directory,
then the result is a \code{config} file in the corresponding \code{"lib"} directory
\code{"\var{...}/lib/\var{app-name}/config"}.
If \code{app:path} is set to a value of the form
\code{"\var{...}/\var{app-name}\opt{.\var{ext}}"},
then the result is a file with a \code{.config} extension in the same directory:
\code{"\var{...}/\var{app-name}.config"}.

If \code{app:path} is not set, the result is a \code{.config} file in the base
directory identified by value of the \code{base-dir} parameter.
This can be useful when loading program code at the REPL during interactive
development.

\defineentry{app:name}
\begin{parameter}
  \code{(app:name \opt{\var{name}})}
\end{parameter}
\hasvalue{} a string or \code{\#f}

The \code{app:name} parameter returns the short descriptive name of the
application, if known.
This value is used by the \code{app-exception-handler} to identify the program
that is reporting an error.
The value is also useful as an argument to \code{display-usage} and
\code{display-help}.

When called with a string, \code{app:name} treats the value as
a file-system path and stores only the last element of the path,
dropping any file extension.

For stand-alone programs, \code{swish-start}
sets this parameter to the path of the running executable.
For scripts or linked programs, \code{swish-start}
sets this parameter to the path of the script or program file.

\defineentry{app:path}
\begin{parameter}
  \code{(app:path \opt{\var{path}})}
\end{parameter}
\hasvalue{} an absolute path or \code{\#f}

The \code{app:path} parameter returns the absolute path to the script or
program, if known.
This value is used by \code{app:config-filename} to determine the
location of the application's configuration file.
For scripts and linked programs, where the actual executable is Swish, this
value is different from that returned by \code{osi\_get\_executable\_path}.

When called with a string, \code{app:path} calls \code{get-real-path} to
obtain an absolute path and stores the result.

\defineentry{base-dir}
\begin{parameter}
  \code{(base-dir \opt{\var{path}})}
\end{parameter}
\hasvalue{} a file-system path

When called without arguments, \code{base-dir} returns the file-system path of
the application's base directory.
Otherwise, the base directory is set to \var{path}, which must specify the
file-system path of a directory that exists and is writable by the application.
Setting \code{base-dir} sets the values of \code{data-dir}, \code{log-file},
\code{tmp-dir}, and \code{web-dir} so that they refer to locations within the
base directory.

When the application starts, \code{base-dir} is set to the current directory.

\defineentry{data-dir}
\begin{parameter}
  \code{(data-dir \opt{\var{path}})}
\end{parameter}
\hasvalue{} a file-system path

The \code{data-dir} parameter holds the file-system path of a directory
in which the application can write persistent data.
Applications should not assume that this directory exists, but should
instead use \code{make-directory-path} when creating files under \code{data-dir}.
Its initial value is the path to a \code{"data"} subdirectory of
the directory specified by \code{base-dir}.

\defineentry{log-file}
\begin{parameter}
  \code{(log-file \opt{\var{filename}})}
\end{parameter}
\hasvalue{} a file name

The \code{log-file} parameter holds the name of the file containing the SQLite
log database managed by the \code{log-db} gen-server, which creates the file
and the requisite directory structure when started.
Its initial value is the path to a \code{"Log.db3"} file in the directory
specified by \code{data-dir}.

\defineentry{tmp-dir}
\begin{parameter}
  \code{(tmp-dir \opt{\var{path}})}
\end{parameter}
\hasvalue{} a file-system path

The \code{tmp-dir} parameter holds the file-system path of a directory
in which the application can write ephemeral data.
Applications should not assume that this directory exists, but should
instead use \code{make-directory-path} when creating files under \code{tmp-dir}.
Its initial value is the path to a \code{"tmp"} subdirectory of
the directory specified by \code{base-dir}.

\defineentry{web-dir}
\begin{parameter}
  \code{(web-dir \opt{\var{path}})}
\end{parameter}
\hasvalue{} a file-system path

The \code{web-dir} parameter holds the file-system path of the directory
used by the HTTP interface (see Chapter~\ref{chap:http}).
Its initial value is the path to a \code{"web"} subdirectory of
the directory specified by \code{base-dir}.

\defineentry{include-line}
\begin{syntax}
  \code{(include-line \var{filenme} \opt{\var{not-found}})}
\end{syntax}
\returns{} see below

The \code{include-line} macro expects a string constant \var{filename}
identifying a file either via absolute path or as a path relative to a
directory in \code{source-directories}.
If such a file can be found at expand time, \code{include-line} expands to a
string containing the first line of that file or the end-of-file object if the
file is empty.
If the file cannot be found, \code{include-line} expands to
\code{(\var{not-found} \var{filename})}, where \var{not-found} defaults to an
expression that raises a syntax error indicating that the file could not be
found.

\defineentry{software-info}
\begin{procedure}
  \code{(software-info)}
\end{procedure}
\hasvalue{} a JSON object

The \code{software-info} procedure returns a JSON object containing the values
stored by
\code{software-product-name},
\code{software-revision}, and
\code{software-version}.
Swish populates these parameters for the
\code{swish} and \code{chezscheme} keys
with values that are determined at compile time.

\defineentry{software-product-name}
\begin{parameter}
  \code{(software-product-name \opt{\var{key}} \opt{\var{value}})}
\end{parameter}
\hasvalue{} a string or \code{\#f}

The \code{software-product-name} procedure stores or retrieves the product
name under the path \code{(\var{key} product-name)} in the
\code{(software-info)} object.
If \var{value} is omitted, this procedure returns the value, if any,
associated with \var{key}, or else \code{\#f}.
If \var{key} and \var{value} are omitted, this procedure returns the value
associated with the key \code{(string->symbol (or (app:name) "swish"))}.
If specified, \var{key} must be a symbol and \var{value} must be a string.
If \var{key} and \var{value} are specified and a value has already been
set for \var{key}, this procedure has no effect.

\defineentry{software-revision}
\begin{parameter}
  \code{(software-revision \opt{\var{key}} \opt{\var{value}})}
\end{parameter}
\hasvalue{} a string or \code{\#f}

The \code{software-revision} procedure stores or retrieves the software
revision under the path \code{(\var{key} revision)} in the
\code{(software-info)} object.
If \var{value} is omitted, this procedure returns the value, if any,
associated with \var{key}, or else \code{\#f}.
If \var{key} and \var{value} are omitted, this procedure returns the value
associated with the key \code{(string->symbol (or (app:name) "swish"))}.
If specified, \var{key} must be a symbol and \var{value} must be a string.
If \var{key} and \var{value} are specified and a value has already been
set for \var{key}, this procedure has no effect.

\defineentry{software-version}
\begin{parameter}
  \code{(software-version \opt{\var{key}} \opt{\var{value}})}
\end{parameter}
\hasvalue{} a string or \code{\#f}

The \code{software-version} procedure stores or retrieves the software
version under the path \code{(\var{key} version)} in the
\code{(software-info)} object.
If \var{value} is omitted, this procedure returns the value, if any,
associated with \var{key}, or else \code{\#f}.
If \var{key} and \var{value} are omitted, this procedure returns the value
associated with the key \code{(string->symbol (or (app:name) "swish"))}.
If specified, \var{key} must be a symbol and \var{value} must be a string.
If \var{key} and \var{value} are specified and a value has already been
set for \var{key}, this procedure has no effect.

\subsection{Program Life Cycle}

\defineentry{app-exception-handler}
\begin{procedure}
  \code{(app-exception-handler \var{obj})}
\end{procedure}
\returns{} unspecified

When a Swish application starts, it sets the default value of the
\code{base-exception-handler} parameter to \code{app-exception-handler}.
This procedure expects a single argument, typically a condition or an object
passed to \code{raise}, which it saves in the parameter \code{debug-condition}.

If \code{app:name} is set, then \code{app-exception-handler} writes a
message to the console error port, prefixed with the application name,
before calling \code{reset}.
If \var{obj} is a condition, then the message is formed by stripping the
``Warning:'' or ``Exception:'' prefix from the output of \code{display-condition}.
Otherwise, the message is generated by calling
\code{exit-reason->english}
on \var{obj}.

If \code{app:name} is not set, then \code{app-exception-handler}
calls the default exception handler.

\defineentry{app-sup-spec}
\begin{parameter}
  \code{(app-sup-spec \opt{\var{start-specs}})}
\end{parameter}
\hasvalue{} a list of child specifications (see page~\pageref{page:child-spec})

The \code{app-sup-spec} parameter supplies the \var{start-specs}
that define the tree of child processes that are started by
\code{app:start} and supervised by the \code{main-sup} gen-server.
The initial value of \code{app-sup-spec} is constructed by calling
\code{make-swish-sup-spec} with \code{swish-event-logger} as the
sole logger.

\defineentry{app:resume}
\begin{syntax}
  \code{(app:resume)}
\end{syntax}

This is an alias for \code{statistics:resume}.

\defineentry{app:shutdown}
\begin{syntax}
  \code{(app:shutdown \opt{\var{exit-code}})}
\end{syntax}

This is an alias for \code{application:shutdown}.

\defineentry{app:start}
\begin{procedure}
  \code{(app:start)}
\end{procedure}
\returns{} \code{ok}

The \code{app:start} procedure calls \code{application:start}
with a \var{starter} thunk that calls \code{supervisor:start\&link}
with the value of the \code{app-sup-spec} parameter.

\defineentry{app:suspend}
\begin{syntax}
  \code{(app:suspend)}
\end{syntax}

This is an alias for \code{statistics:suspend}.

\defineentry{make-swish-sup-spec}
\begin{procedure}
  \code{(make-swish-sup-spec \var{loggers})}
\end{procedure}
\returns{} a list of child specifications (see page~\pageref{page:child-spec})

The \code{make-swish-sup-spec} procedure builds a list of child specifications
representing the default supervision tree described in
Section~\ref{sec:default-supervision-tree}.
The \var{loggers} argument is a list of \code{<event-logger>} tuples that will
be passed to \code{log-db:setup} when the \code{log-db} gen-server is started.

\defineentry{swish-start}
\begin{procedure}
  \code{(swish-start \var{arg} \ldots)}
\end{procedure}
\returns{} see below

When a stand-alone program starts, it initializes several parameters:
\code{app:name}, \code{app:path}, \code{command-line}, and
\code{command-line-arguments}.
When a Swish script, linked program, or REPL is started, it performs this
initialization by applying \code{swish-start} to the original command-line
arguments.

The \code{swish-start} procedure expects zero or more strings as arguments
and scans them left to right.
It adds each string that begins with a single dash (\sopt{}) or double dash
(\lopt{}) to a set of options to be handled later.
If it encounters the string \str{\lopt{}}, it sets the REPL option and
saves the remaining arguments as files to be loaded.
If it encounters a string that does not begin with a dash, it stops scanning
the arguments and marks that string and those that follow as ordinary
arguments.

If \code{swish-start} finds \lopt{help} among the options,
it displays a summary of Swish's command-line options and returns.
If it finds \lopt{version} among the options,
it displays version information and returns.
If the REPL option is set or it finds no ordinary arguments, then it
suppresses the startup message and waiter prompt 
if \lopt{quiet} is among the options,
then loads any files that were specified,
and then calls \code{new-cafe}. 
Otherwise, it treats the first ordinary argument as a script or linked program.
In this case, it installs the ordinary arguments as
the value of \code{command-line},
sets \code{app:name} and \code{app:path} to the first of them,
sets \code{command-line-arguments} to the remaining ones,
and clears \code{app:config}
before loading the script or linked program.
If an error occurs while loading the script or linked program,
Swish exits with an exit code of 1.
Otherwise, Swish exits normally.

When it starts a REPL, script, or linked program, \code{swish-start} attempts
to import any libraries found in \code{library-list} whose library path begins
with the symbol \code{swish}.

\subsection{Foreign Interface}

Scheme libraries that rely on shared objects must arrange
to call \code{load-shared-object} before calling \code{foreign-procedure} or
calling \code{make-ftype-pointer} on a function ftype.
We must consider shared-object naming conventions and search paths on
different platforms when calling \code{load-shared-object}.
We can provide its filename argument through some combination of conditional
compilation, hard-coded relative or absolute paths, and general computation.

The \code{provide-shared-object} and \code{require-shared-object} procedures
in this section offer a simple way to:
factor platform dependencies out of client code,
load shared objects by absolute path to avoid search,
specify shared object file names as paths relative
to the application configuration file, and
hook the operation that loads shared object code.

These procedures coordinate via the application configuration object stored in
\code{app:config}, which is populated on demand by reading the file specified
by \code{app:config-filename}.
A Scheme library can call \code{require-shared-object} before its
\code{foreign-procedure} definitions and rely on demand-loading of the
configuration file to supply the absolute path to the shared object code when
the library is invoked.
This simplifies a process that can otherwise be complicated
by the fact that Chez Scheme invokes a library lazily as soon as
one of its exports may be referenced.

To see how these procedures are decoupled, consider the following REPL
transcript on a 64-bit Linux machine:
\codebegin
Swish Version 0.0.0
> (provide-shared-object 'libc "/lib64/libc.so.6")
> (foreign-entry? "fork")
#f
> (require-shared-object 'libc)
> (foreign-entry? "fork")
#t
> (let ([op (open-file-to-write (app:config-filename))])
    (on-exit (close-port op)
      (json:write op (app:config) 0)))
> (exit)
\codeend

If we start a new Swish REPL, we can call \code{require-shared-object}
without first calling \code{provide-shared-object}
because we explicitly wrote the value of \code{app:config} to the location
specified by \code{app:config-filename} before exiting the original REPL.
\codebegin
Swish Version 0.0.0
> (foreign-entry? "fork")
#f
> (require-shared-object "libc")
> (foreign-entry? "fork")
#t
> (display (utf8->string (read-file (app:config-filename))))
\{
  "swish": \{
    "shared-object": \{
      "libc": \{
        "a6le": \{
          "file": "/lib64/libc.so.6"
        \}
      \}
    \}
  \}
\}
\codeend

\defineentry{provide-shared-object}
\begin{procedure}
  \code{(provide-shared-object \var{so-name} \var{filename})}
\end{procedure}
\returns{} unspecified

The \code{provide-shared-object} procedure expects a symbol
\var{so-name} as the generic name of a shared object and
a string \var{filename} that is the
name of the actual file containing the shared object code.
This procedure records \var{filename} in \code{app:config} under a
path that includes \code{so-name} and the host \code{machine-type}.

\defineentry{require-shared-object}
\begin{procedure}
  \code{(require-shared-object \var{so-name} \opt{\var{handler}})}
\end{procedure}
\returns{} unspecified

The \code{require-shared-object} procedure loads a shared object via an
absolute path to prevent \code{load-shared-object} from searching for the
shared object file in a system-specific set of directories.

The \var{so-name} argument is a symbol that specifies the generic name of a
shared object.
The optional \var{handler} is a procedure that expects three arguments:
\var{filename} is the absolute path to the shared object file,
\var{key} is \var{so-name}, and \var{dict} is the hashtable
retrieved from \code{app:config} via a path that includes
\var{so-name} and the host \code{machine-type}.
This hashtable contains a key \code{file} whose value
is the file name originally supplied by \code{provide-shared-object}.
The handler might examine other keys within the hashtable before
loading the shared object.
The default handler simply calls \code{load-shared-object} on \var{filename}.

If the file name retrieved from \code{app:config} is a relative path,
then \code{require-shared-object} determines the absolute path by
treating the file name as a path relative to the parent directory
containing the application configuration file name returned by \code{app:config-filename}.

\defineentry{define-foreign}
\begin{syntax}
  \code{(define-foreign \var{name} (\var{arg-name} \var{arg-type}) ...)}
\end{syntax}
\expandsto{} \antipar\codebegin
(begin
  (define \var{name}* (foreign-procedure (symbol->string '\var{name}) (\var{arg-type} ...) ptr))
  (define (\var{name} \var{arg-name} ...)
    (match (\var{name}* \var{arg-name} ...)
      [(,who . ,err)
       (guard (symbol? who))
       (io-error '\var{name} who err)]
      [,x x])))
\codeend

The \code{define-foreign} macro defines two procedures: \code{\var{name}*}
is a raw foreign procedure that expects the specified argument types and returns
a \code{ptr} value, while \var{name} is a wrapper that calls \code{\var{name}*}
and raises an \code{io-error} if it returns an error pair.

\input{swish/testing.tex}
